Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-01-07 paper/tidy-missing-data-paper.Rmd
Restructuring headings of the paper

---
documentclass: jss
author:
  - name: Nicholas Tierney
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics
      E774, Menzies Building
      Monash University, Clayton
    email: \email{nicholas.tierney@gmail.com}
    url: http://www.njtierney.com
  - name: Dianne Cook
    affiliation: Monsh University
title:
  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations with the Package \\pkg{naniar}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations with the Package naniar"
  # For running headers, if needed
  short:     "\\pkg{naniar}: Explore missing data"
abstract: >
  Despite the large body of research on missing value distributions and imputation, there is comparatively little literature with a focus on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as an integral part of data analysis workflows. We define both new data structures and new verbs (functions) to perform common operations. Together, these provide a cohesive framework for handling, exploring, and imputing missing values. These methods have been made available in the R package `naniar`.
keywords:
  # at least one keyword must be supplied
  formatted: [workflow, statistical computing, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
  plain:     [workflow, statistical computing, data science, data visualization, tidyverse, data pipeline, R]
preamble: >
  \usepackage{amsmath}
  \usepackage[english]{babel}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    includes:
      in_header: header.tex
    keep_tex: yes
    citation_package: 'natbib'
bibliography: tidy-missing-data-paper.bib
# csl: jcgs.csl
---

@@

@@

## Exploration {#exploration}

@@

Applying tidyverse principles to the domain of missing data clarifies and
facilitates missing data exploration, visualization, and imputation. This section
discusses how these principles are applied, and proceeds in three parts. Section
\@ref(data-structure) discusses _nabular_ data, a tidy data structure for
missing values that tracks missings throughout the analysis and allows for
special missing values. Section \@ref(vis-summaries) covers visualizations to
understand missingness. Section \@ref(num-sum) discusses data summaries, and
Section \@ref(verbs) covers the common "verbs" used to solve problems
when working with missing data. The design and naming of these features is an
important component that helps the analyst compose them efficiently in data pipelines, and is discussed throughout.

@@

(**consider removing these longer expansions of these points?**)

@@

## Visual summaries {#vis-summaries}

Visualizing missingness in a dataset is essential to understand its structure. This section discusses easy to remember visualizations that compose with other functions in a pipeline. Four different areas of visualization of missing data are discussed: overviews (Section \@ref(overview-plots)); univariate (Section \@ref(univariate-plots)), bivariate (Section \@ref(bivariate-plots)), and multivariate (Section \@ref(multivariate-plots)). All plots are created using \pkg{ggplot2}, giving users clearer control over the plot appearance, and customizability.

### Overview plots {#overview-plots}

@@

### Univariate plots {#univariate-plots}

@@

#### Imputing values below the range {#imp-below-range}

@@

#### Univariate split by missingness {#vis-split-by-miss}

@@

### Bivariate plots {#bivariate-plots}

@@

### Multivariate plots {#multivariate-plots}

@@

## Numerical summaries {#num-sum}

@@

### Single number summaries {#single-num-sum}

@@

### Summaries and tabulations of missing data {#sum-tab-missings}

@@

### Combining numerical summaries with grouping operations {#num-sum-w-group}

>>>>

---
documentclass: jss
author:
  - name: Nicholas Tierney
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics
      E774, Menzies Building
      Monash University, Clayton
    email: \email{nicholas.tierney@gmail.com}
    url: http://www.njtierney.com
  - name: Dianne Cook
    affiliation: Monsh University
title:
  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # If you use tex in the formatted title, also supply version without
  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # For running headers, if needed
  short:     "Explore missingness with \\pkg{naniar}"
abstract: >
  Despite the large body of research on missing value distributions and imputation, there is comparatively little literature with a focus on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as an integral part of data analysis workflows. We define both new data structures and new verbs (functions) to perform common operations. Together, these provide a cohesive framework for handling, exploring, and imputing missing values. These methods have been made available in the R package `naniar`.
keywords:
  # at least one keyword must be supplied
  formatted: [workflow, statistical computing, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
  plain:     [workflow, statistical computing, data science, data visualization, tidyverse, data pipeline, R]
preamble: >
  \usepackage{amsmath}
  \usepackage[english]{babel}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    includes:
      in_header: header.tex
    keep_tex: yes
    citation_package: 'natbib'
bibliography: tidy-missing-data-paper.bib
# csl: jcgs.csl
---

@@

# Existing Software {#existing-software}

@@

### Exploration {#exploration}

@@

Applying tidyverse principles to the domain of missing data clarifies and
facilitates missing data exploration, visualization, and imputation. This section
discusses how these principles are applied, and proceeds in three parts. Section
\@ref(data-structure) discusses _nabular_ data, a tidy data structure for
missing values that tracks missings throughout the analysis and allows for
special missing values. Section \@ref(graphics) covers visualizations to
understand missingness. Section \@ref(num-sum) discusses data summaries, and
Section \@ref(verbs) covers the common "verbs" used to solve problems
when working with missing data. The design and naming of these features is an
important component that helps the analyst compose them efficiently in data pipelines, and is discussed throughout.

@@

@@

# Graphics {#graphics}

Visualizing missingness in a dataset is essential to understand its structure. This section discusses easy to remember visualizations that compose with other functions in a pipeline. Four different areas of visualization of missing data are discussed: overviews (Section \@ref(overview-graphics)); univariate (Section \@ref(univariate-graphics)), bivariate (Section \@ref(bivariate-graphics)), and multivariate (Section \@ref(multivariate-graphics)). All plots are created using \pkg{ggplot2}, giving users clearer control over the plot appearance, and customizability.

## Overview graphics {#overview-graphics}

@@

## Univariate graphics {#univariate-graphics}

@@

### Imputing values below the range {#imp-below-range}

@@

### Univariate split by missingness {#vis-split-by-miss}

@@

## Bivariate graphics {#bivariate-graphics}

@@

## Multivariate graphics {#multivariate-graphics}

@@

# Numerical summaries {#num-sum}

@@

## Single number summaries {#single-num-sum}

@@

## Summaries and tabulations of missing data {#sum-tab-missings}

@@

## Combining numerical summaries with grouping operations {#num-sum-w-group}
